---
title: "DASC 500: Introduction to Data Analytics"
subtitle: "Clustering Algorithms"
date: '`r format(Sys.Date(), "%d %B %Y")`'
footer: "`r paste(rmarkdown::metadata$title, ' - ',rmarkdown::metadata$subtitle)`"
author: "Maj Jason Freels PhD"
output: 
  slidy_presentation:
    smart: no
    fig_caption: yes
graphics: yes
---

# Overview

<style>
div.plotly.html-widget {width: 800px;
                   height: 800px;
                   margin-left:auto;
                   margin-right:auto;}
</style>

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      fig.align = 'center',
                      message = F,
                      warning = F)

shiny::includeCSS("../resources/css/flatly-style.css")
shiny::includeCSS("../resources/css/r-title.css")
shiny::includeScript("../resources/js/jquery.min.js")
shiny::includeScript("../resources/js/jkf-scroll.js")

library(knitr)
library(reticulate)
library(rprojroot)
knitr::opts_chunk$set(echo = TRUE, message = F, comment = NA, warning = F)
root = find_root(is_git_root)
```

```{python, echo=F}
import warnings
warnings.filterwarnings('ignore')
```

## Introduction

- Clustering algorithms are a family of methods used to identify subgroups within a data set by classifying observations into similar categories 

- We implement clustering algorithms because we want to partition the observations in such a way to help guide further analyses

- Since there isn't a response variable, clustering is an example of **unsupervised** learning algorithms

- In this lecture we'll look at two common types of clustering algorithms

    + K-Means clustering
    + Hierarchical clustering

- We’ll demonstrate these methods using the `Mall_Customers` data set which can be accessed from <a target=" "href="https://raw.githubusercontent.com/SteffiPeTaffy/machineLearningAZ/master/Machine%20Learning%20A-Z%20Template%20Folder/Part%204%20-%20Clustering/Section%2024%20-%20K-Means%20Clustering/Mall_Customers.csv">**this site**</a> 


- This data set contains 5 columns 

    + `CustomerID`: Numeric identifier for each customer in the data set
    + `Genre`: Character string indicating the customer's gender
    + `Annual Income (k$)`: The customer's annual income (in thousands of dollars) 
    + `Age`: Numeric variable of the customer's age in years
    + `Spending Score (1-100)`: Numeric indicator of the customer's spending habits

## Setting up the workspace

- To carry out this analysis you'll need to import the following libraries and modules 

```{python}
# for basic mathematics operation 
import numpy as np
import pandas as pd
from pandas import plotting

# for visualizations
import matplotlib.pyplot as plt
import seaborn as sb
plt.style.use('fivethirtyeight')
fig = plt.gcf() 
fig.set_size_inches(12, 8)
```

- With the packages imported, let's read in the data from the URL provided above

```{python}
# importing the data set
data_url = "https://raw.githubusercontent.com/SteffiPeTaffy/machineLearningAZ/master/Machine%20Learning%20A-Z%20Template%20Folder/Part%204%20-%20Clustering/Section%2024%20-%20K-Means%20Clustering/Mall_Customers.csv"

df = pd.read_csv(data_url)
```

```{r echo=FALSE}
knitr::kable(head(py["df"],20))
```

- This data set has been partially cleaned to simplify our analysis, but in general the data should be prepared as follows:

    1. Ensure that each row contains one observation
    2. Ensure that each column contains a single variable
    3. Ensure that any missing values in the data are removed or can be estimated
    4. Ensure that the data are standardized to make variables comparable (this means that the variables are transformed such that they have a mean of zero and a standard deviation of one

- For step 3 we can check that there aren't any "troublesome" values by invoking the `isnull()` function over the columns and rows in the data set

```{python}
# checking if there is any NULL data

df.isnull().any().any()
```

- For step 4, the data have not been standardized - let's do that now 

- For each column in our data set that contain numeric values we

    + Subtract from each value the mean of that column
    + Divide this result by the standard deviation of that column 

```{python}
df_norm = df

df_norm.iloc[:,2] = (df.iloc[:,2] - df.iloc[:,2].mean()) / df.iloc[:,2].std()
df_norm.iloc[:,3] = (df.iloc[:,3] - df.iloc[:,3].mean()) / df.iloc[:,3].std()
df_norm.iloc[:,4] = (df.iloc[:,4] - df.iloc[:,4].mean()) / df.iloc[:,4].std()

df_norm
```

- With the data prepared let's dive in to clustering algorithms

# Clustering overview

## Distance Measures

- Clustering algorithms are implemented on at least two components (columns) in our data

    + The first component can be thought of as the `x` coordinate in an `x-y` plane
    + The second component can be thought of as the `y` coordinate in an `x-y` plane 

- It's important to understand that clustering algorithms pre-suppose a method used to compute the distance or the dissimilarity between each pair of observations

- The result of this computation is a dissimilarity or distance matrix

    + There are many methods to calculate this distance information
    + The choice of distance measures is a critical step in clustering and defines how the similarity of two elements (x, y) is calculated and it will influence the shape of the clusters

- Two classical methods for distance measures are *Euclidean* and *Manhattan distances*, which are defined as:

    + __Euclidean distance:__ $d_{euc}(x,y) = \sqrt{\sum^n_{i=1}(x_i - y_i)^2}$
    + __Manhattan distance:__ $d_{man}(x,y) = \sum^n_{i=1}|(x_i - y_i)|$

- Other dissimilarity measures exist such as correlation-based distances

- Correlation-based distance is defined by subtracting the correlation coefficient from $1$ 

- There are several different methods for computing correlation-based distances, such as:

    + __Pearson correlation distance__
    + __Spearman correlation distance__
    + __Kendall correlation distance__

- The choice of distance measures is very important, as it has a strong influence on the clustering results

- For many software tools, the default distance measure is the Euclidean distance

- However, depending on the type of the data and the research questions, other dissimilarity measures might be preferred and you should be aware of the options.

- In Python we can create the dissimilarity matrix using the functions <a target=" " href="https://docs.scipy.org/doc/scipy/reference/generated/scipy.spatial.distance.pdist.html">pdist()</a> and <a target=" " href="https://docs.scipy.org/doc/scipy/reference/generated/scipy.spatial.distance.squareform.html#scipy.spatial.distance.squareform">squareform()</a>

```{python}
from scipy.spatial.distance import squareform
from scipy.spatial.distance import pdist

dm = pd.DataFrame(squareform(pdist(df_norm.iloc[:, 2:])))

dm
```

## Clustering Algorithms

- __K-means clustering__ is a relatively simple method used to cluster observations into a pre-specified set of `k` groups

- __Hierarchical clustering__ is an alternative approach to k-means that does not require us to pre-specify the number of clusters 

- These two methods are discussed further in the following sections

```{r child="kmeans.Rmd", eval=!FALSE}
```


```{r child="heirarchical.Rmd", eval=!FALSE}
```


