---
title: "DASC 500: Introduction to Data Analytics"
subtitle: "An Introduction to Clustering Algorithms"
date: '`r format(Sys.Date(), "%d %B %Y")`'
footer: "DASC 500 (Introduction Supervised Learning Algorithms)"
author: "Maj Jason Freels PhD"
output: 
  slidy_presentation:
    smart: no
    fig_caption: yes
graphics: yes
---

# Overview

<style>
div.plotly.html-widget {width: 800px;
                   height: 800px;
                   margin-left:auto;
                   margin-right:auto;}
</style>

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      fig.align = 'center',
                      message = F,
                      warning = F)

shiny::includeCSS("../resources/css/flatly-style.css")
shiny::includeCSS("../resources/css/r-title.css")
shiny::includeScript("../resources/js/jquery.min.js")
shiny::includeScript("../resources/js/jkf-scroll.js")

library(knitr)
library(reticulate)
library(rprojroot)
knitr::opts_chunk$set(echo = TRUE, message = F, comment = NA, warning = F)
root = find_root(is_git_root)
```


 Because there isn't a response variable, this is an unsupervised method, which implies that it seeks to find relationships between the $$n$$ observations without being trained by a response variable. Clustering allows us to identify which observations are alike, and potentially categorize them therein.

- Clustering algorithms are a family of methods for identifying subgroups within a data set

- Clustering methods help classify observations into similar categories 

    + Categories of customers
    + Categories of students

- We want observations in the same group to be similar and observations in different groups to be dissimilar

To perform a cluster analysis, generally, the data should be prepared as follows:

1. Rows are observations (individuals)
2. columns are variables
2. Any missing value in the data must be removed or estimated.
3. The data must be standardized (i.e., scaled) to make variables comparable. Recall that, standardization consists of transforming the variables such that they have mean zero and standard deviation one.[^scale] 

Here, we’ll use `Mall_Customers`, data set which contains statistics in arrests per 100,000 residents for assault, murder, and rape in each of the 50 US states in 1973. 

- The data set may be accessed from <a target=" " href="https://raw.githubusercontent.com/SteffiPeTaffy/machineLearningAZ/master/Machine%20Learning%20A-Z%20Template%20Folder/Part%204%20-%20Clustering/Section%2024%20-%20K-Means%20Clustering/Mall_Customers.csv">**this site**</a>

## Setting up our workspace

```{python}
# for basic mathematics operation 
import numpy as np
import pandas as pd
from pandas import plotting

# for visualizations
import matplotlib.pyplot as plt
import seaborn as sns
plt.style.use('fivethirtyeight')
```

```{python, eval=!F}
# for interactive visualizations
import plotly.offline as py
from plotly.offline import init_notebook_mode, iplot
import plotly.graph_objs as go
from plotly import tools
#init_notebook_mode(connected = True)
import plotly.figure_factory as ff
```

```{python}
# importing the dataset
data_url = "https://raw.githubusercontent.com/SteffiPeTaffy/machineLearningAZ/master/Machine%20Learning%20A-Z%20Template%20Folder/Part%204%20-%20Clustering/Section%2024%20-%20K-Means%20Clustering/Mall_Customers.csv"

data = pd.read_csv(data_url)
```

```{r echo=FALSE}
DT::datatable(py["data"])
```


```{python}
# describing the data

desc = ff.create_table(data.describe())
py.iplot(desc)
```

```{python}
# checking if there is any NULL data

data.isnull().any().any()
```


```{r}

```


## Clustering Algorithms 

- K-means clustering a simple and commonly used clustering method for splitting a data set into `k` groups

- Hierarchical clustering is an alternative approach to k-means clustering for identifying groups in the data set. It does not require us to pre-specify the number of clusters to be generated as is required by the k-means approach. Furthermore, hierarchical clustering has an added advantage over K-means clustering in that it results in an attractive tree-based representation of the observations, called a dendrogram.

```{r child="kmeans.Rmd", eval=FALSE}
```


```{r child="heirarchical.Rmd", eval=FALSE}
```