---
title: "DASC 500: Introduction to Data Analytics"
subtitle: "Statistical Inference (Part 2 - Applications)"
date: '`r format(Sys.Date(), "%d %B %Y")`'
footer: "DASC 500 (Introduction to Statistical Inference)"
author: "Maj Jason Freels PhD"
output: 
  slidy_presentation:
    smart: no
    fig_caption: yes
graphics: yes
---

# Overview

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.align = 'center')
shiny::includeCSS("../resources/css/flatly-style.css")
shiny::includeCSS("../resources/css/r-title.css")
shiny::includeScript("../resources/js/jquery.min.js")
shiny::includeScript("../resources/js/jkf-scroll.js")

library(knitr)
library(reticulate)
library(rprojroot)
knitr::knit_engines$set(python = reticulate::eng_python)
knitr::opts_chunk$set(echo = TRUE, message = F, comment = NA, warning = F)
root = find_root(is_git_root)
```

## In this presentation

- Estimation

    + Point estimators for the mean
    + Point estimators for the variance
    + Constructing confidence intervals 

- Hypothesis tests

    + Two sided hypothesis test for the mean
    + One-sided hypothesis test for the mean

## Setup

- To follow along with this you'll need to import the following libraries into your Python workspace (or Google Colab instance)

```{python}
import statistics as st
import numpy as np
import seaborn as sb
import matplotlib.pyplot as plt
import pandas as pd
import scipy as sp
```

```{r child='estimation2.Rmd', eval=!FALSE}
```

```{r child='confidence2.Rmd', eval=!FALSE}
```

```{r child='hypothesis3.Rmd', eval=!FALSE}
```